Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-24 manuscript.Rmd
Implications Cultural Sociology I

The second model summarized by Kiley and Vaisey [-@kiley2020], an “Active Updating Model” (AUM), posits that people continually update their personal culture as they move through life. This model suggests people change their personal culture as they adapt and make new meanings when encountering new social environments, discourses, and information [@gross2009; @swidler2001]. This model underlies, among others, theories of cultural diffusion [@christakis2010], attitude alignment [@dellaposta2020], and polarization [@bail2018] It is also implicit in most studies that ask whether specific experiences, changes in social roles, or political events, affect personal culture [@gelman2021; @visser2004; @slothuus2021].

@@

## A Case Study: College Education

@@

So far, the debate in cultural sociology has centered around the presence and significance of cultural change in adults. Yet, arguing solely over its existence sets the wrong theoretical focus. In any adult population, some degree of personal cultural change is inevitable. Likewise, no viable theoretical perspective would expect just one process --stable differences or internal change between people-- to characterize the differences in personal cultural observed between them. Our results underscore this perspective, showing a combination of interpersonal differences and intrapersonal change across all items. [DELETED SENTECE, MISSPLACED]

A more reasonable theoretical debate pushed by previous work is the relative importance of these two processes for explaining cultural variation. It is difficult to draw a single conclusion on this question across a wide range of questions measured on different scales and over different durations of time, but the overall picture, with important caveats, suggests that intrapersonal change during adulthood is not the dominant explanation for cultural differences among adults for most items we measure in surveys. There simply is not enough evidence of intrapersonal change on these items relative to the amount of difference we observe across people.

To be clear, none of our results should be interpreted as suggesting that adults do not change or are incapable of changing. On all questions, some members of the sample change their personal culture at least a little bit. Our results also should not be taken to imply that intrapersonal change is not important or consequential. Small numbers of changers can have important consequences in the aggregate. A shift from 49 percent supporting gay marriage to 51 percent supporting it can have significant political ramifications, even as most respondents remain fixed in their preferences. And our results should not detract from work trying to estimate the mechanisms through which personal cultural differences --however small-- emerge among adults. Just because change is small does not mean the mechanisms that produce it are not important to understand.

Our results should be read principally as suggesting that if researchers want to explain why people differ in their personal culture, it is imperative to look at the conditions in which they were raised and the experiences that happened before they reached adulthood. To the extent that we want to explain what makes adults different, focusing attention on peoples’ transition between social roles, their selection into and out of organizational environments, and changes in their social networks, as much sociological work does, appears to explain a much smaller proportion of their differences than the formative experiences that happen early in life.

In this way, our findings echo a range of recent causal inference work that suggests selection effects, rather than treatment effects, seem to explain the majority of differences in personal culture between people in different social roles or positions [@campbell2016; @wodtke2018]. While some people clearly change over time as they move into new social roles or into new social environments, this amount of change does not appear to be sufficient to explain what are often substantial differences between people in different roles. Our results suggest that, to the extent we observe differences in personal culture between people in different social roles, such as parenthood, having children, education, or working in positions of authority [@longest2013], or between people in different occupations [@weeden2005], our prior expectation should be that this difference is selection, rather than treatment.

>>>>

The second model summarized by Kiley and Vaisey [-@kiley2020], an “Active Updating Model” (AUM), posits that people continually update their personal culture as they move through life. This model suggests people change their personal culture as they adapt and make new meanings when encountering new social environments, discourses, and information [@gross2009; @swidler2001]. This model underlies, among others, theories of cultural diffusion [@christakis2010], attitude alignment [@dellaposta2020], and polarization [@bail2018]. It is also implicit in most studies that ask whether specific experiences, changes in social roles, or political events, affect personal culture [@gelman2021; @visser2004; @slothuus2021].

@@

## A Case Study — The Role of College for Change in Political Dispositions Among Adults

@@

So far, sociologists interested in understanding cultural differences have largely asked about the presence and significance of cultural change in adults [e.g., @kiley2020; @vaisey2021; @lersch2023]. Yet doing so, might have unnecessarily limited and even set the wrong theoretical focus. In any adult population, some degree of cultural change in people is inevitable. Likewise, no viable theoretical perspective would expect just one process -- stable differences or internal change between people -- to underlie the differences in personal culture observed between them. Our results reinforce this point, showing the relevance of both interpersonal differences and intrapersonal change in understanding differences in personal culture across all items analyzed. [TRIED TO BETTER TUCK IT IN, BUT COULD LET GO OF THE LAST SENTENCE]

Recent findings suggest that it is theoretically more insightful to ask about the relative importance of these two processes in understanding cultural variation. Drawing a unified conclusion from data on various cultural forms, measured on different scales and across different time frames is challenging. However, the general pattern suggests that, for most survey items, intrapersonal change in adulthood is not the primary reason for cultural differences among people.

To clarify, our findings do not imply that adults remain static or that changes are inconsequential. Even minor shifts in personal culture, such as a 2% change in support for gay marriage, can have significant ramifications. While the majority may remain consistent in their views, understanding the underlying mechanisms of even such minor shifts remains a crucial task for the sociology of culture [COULD SHORTEN TO "EVEN SUCH MINOR CULTURAL SHIFTS REMAIN CRUCIAL" BUT WANTED TO APPEAL TO THE SCHOLARS IN THE FIELD...].

Our findings primarily suggest that understanding variations in personal culture requires scholars to examine the conditions and experiences of early life. While transitions between social roles, the experience of organizational environments [CAN WE SAY SOCIAL INSTITUTIONS?], and changes in social networks are often emphasized in sociological research, these factors seem to account for a smaller proportion of differences among adults than the formative experiences early in life.

This IMPLICATION/CONCLUSION aligns with a range of recent causal inference work suggesting that selection effects, rather than treatment effects, predominantly account for personal cultural differences among individuals in varied social roles and positions [@campbell2016; @wodtke2018]. While some individuals clearly change as they transition into new roles or environments, this magnitude of this change seems to be insufficient to explain the pronounced differences among people in diverse roles. When observing differences in personal culture across social roles, such as parenthood, education, or professional authority [@longest2013], or across occupations [@weeden2005], our initial presumption should lean towards selection as their primary cause, rather than treatment.
